---
title: "Does age impact quality of life in patients undergoing surgery for degenerative spinal diseases?"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
  html_notebook: 
    df_print: default
    toc: yes
    code_folding: hide
    theme: readable
    number_sections: yes
---
---

**Data introduction** 

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)

setwd("C:/Users/pedro/OneDrive/Ambiente de Trabalho/Artigo Qualidade de Vida Marisa")
data <- read_csv2("eq5d.csv")

```
---
 
**Table 1**
```{r echo=TRUE, message=FALSE, warning=FALSE, results='asis'}
#transform variables in factors with labels
data_fact2 <- data

data_fact2$Gender <- factor(data_fact2$Gender, 
                            levels = c(0, 1), labels = c("Female","Male"))

data_fact2 <- data_fact2 %>% mutate(Age=(if_else(Age>64, "Age \U2265 65", "Age < 65")))
data_fact2$Age <- as.factor(data_fact2$Age)

data_fact2$BMI <- 
  factor(data_fact2$BMI, 
         levels = c(1, 2,3), labels = c("<25","25-30", ">30"))

data_fact2$Education3 <- 
  factor(data_fact2$Education3, 
         levels = c(1, 2,3), labels = c(" \U2264 4y","6-12y", ">12y"))

data_fact2$Symptoms_before_Cx <- 
  factor(data_fact2$Symptoms_before_Cx,
         levels = c(1, 2,3), labels = c("<3 m","3-12 m", ">12 m"))

data_fact2$Type_Cx <- 
  factor(data_fact2$Type_Cx, 
         levels = c(1, 2,3), 
         labels = c("Cervical", "Lumbar non fusion", "Lumbar  fusion"))

data_fact2$Levels_Cx_number <- 
  factor(data_fact2$Levels_Cx_number,
         levels = c(1, 2,3), labels = c("1", "2", ">=3"))

dico <- c(9,19, 11:13)
data_fact2[, dico] <- lapply(data_fact2[, dico], factor, 
                             levels = c(0, 1), labels = c("no","yes"))

library(arsenal)
library(kableExtra)

variables <- c(2:5, 9, 11:14, 17:20, 21)
data_fact3 <- data_fact2[,variables]

f1 <- formulize(x=c(1,3,4,6:14), data = data_fact3)

#Use tableby from arsenal, missing values were removed in "cat.stats" (open in global environment to understand)
#f1 contains the variables

tab_1a <- tableby(f1, data = data_fact3, 
                  control = tableby.control(cat.stats = "countpct")) 


summary(tab_1a)
```

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='asis'}
f2 <- formulize(y=2, x=c(1,3,4, 6:14), data = data_fact3)
tab_1b <- tableby(f2, data=data_fact3, 
                  control = tableby.control(numeric.stats = "meansd", 
                                            test = TRUE, 
                                            total = FALSE, 
                                            cat.stats = "countpct"))

summary(tab_1b, pfootnote=TRUE)
```
 
 
---

**Preoperative QoL decreases with age**

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
cor <- cor.test(data$Age, data$EQ5D_Pre)
c(cor$p.value, cor$estimate)
```

  
---

**Diverging stacked bar charts, sorted by status**  
(Credits: https://rpubs.com/tskam/likert  
https://stackoverflow.com/questions/37761096)  


```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

library(HH)

#transform categorical variables in factors
data_fact <- data
factors <- c(1,2,4:19,22:26, 28:34)
data_fact[factors] <- lapply(data_fact[factors], as.factor)

#Making the table long
survey <- gather(data_fact, Domain, Response, c(22:26,28:32))

survey$Domain <- as.factor(survey$Domain)
survey$Response <- as.factor(survey$Response)
survey_df <- table(survey$Domain,survey$Response) %>% as.data.frame.matrix()
colnames(survey_df) <- c("No problems","Some Problems","Extreme problems")

#reorder rows to EQ5D standard and pre-post surgery comparison
survey_df <- survey_df[c(8, 3, 10, 5, 6, 1, 9, 4, 7, 2), ]

#dataframe row names are converted into a column, columns are reordered, for better usage
survey_df <- tibble::rownames_to_column(survey_df, var="Domain")
survey_df <- survey_df[ , c(1,4,3,2)]


#basic diverging stacked bar chart with percentages as labels

##define a custom panel function (to show and align percentage values) 

origNames =  colnames(survey_df)

myPanelFunc <- function(...){
  panel.likert(...)
  vals <- list(...)
  DF <- data.frame(x=vals$x, y=vals$y, groups=vals$groups)

  ### some convoluted calculations here...
  grps <- as.character(DF$groups)
  for(i in 1:length(origNames)){
    grps <- sub(paste0('^',origNames[i]),i,grps)
  }

  DF <- DF[order(DF$y,grps),]

  DF$correctX <- ave(DF$x,DF$y,FUN=function(x){
    x[x < 0] <- rev(cumsum(rev(x[x < 0]))) - x[x < 0]/2
    x[x > 0] <- cumsum(x[x > 0]) - x[x > 0]/2
    return(x)
  })

  subs <- sub(' Positive$','',DF$groups)
  collapse <- subs[-1] == subs[-length(subs)] & DF$y[-1] == DF$y[-length(DF$y)]
  DF$abs <- abs(DF$x)
  DF$abs[c(collapse,FALSE)] <- DF$abs[c(collapse,FALSE)] + DF$abs[c(FALSE,collapse)]
  DF$correctX[c(collapse,FALSE)] <- 0
  DF <- DF[c(TRUE,!collapse),]

  DF$perc <- round(ave(DF$abs,DF$y,FUN=function(x){x/sum(x) * 100}),0)

  panel.text(x=DF$correctX, y=DF$y, label=paste0(DF$perc,'%'), cex=0.7)
}

##stacked bar chart

likert(Domain ~., survey_df, as.percent=TRUE, 
       positive.order=TRUE, rightAxis=FALSE, 
       main = "EQ-5D domains - pre vs postoperative, sorted by status",
       xlab="Percentage",
       panel=myPanelFunc)

```
  
  
**Diverging stacked bar charts, separation by domain**

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

#stacked bar chart - separation by question
Question <-c("1","1","2","2","3","3","4","4","5","5")
survey_df2<-cbind(survey_df,Question)

likert(Domain ~ .| Question, data=survey_df2,
       layout=c(1,5),
       scales=list(y=list(relation="free")),
       between=list(y=1),
       strip.left=strip.custom(bg="gray97"),
       par.strip.text=list(cex=1.1, lines=2),
       strip=FALSE,
       cex=1.2,
       as.percent=TRUE, 
       positive.order=FALSE,
       rightAxis=FALSE,
       main = "EQ-5D domains - pre vs postoperative",
       xlab="Percentage",
       resize.height.tuning=1)
```
  
  
**Diverging stacked bar charts, sorted by status, by age group**  

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

oldest <- filter(survey, Age>=65)
youngest <- filter(survey, Age<65)

survey_df_o <- table(oldest$Domain,oldest$Response) %>% as.data.frame.matrix()
survey_df_y <- table(youngest$Domain,youngest$Response) %>% as.data.frame.matrix()



colnames(survey_df_o) <- c("No problems","Some Problems","Extreme problems")
colnames(survey_df_y) <- c("No problems","Some Problems","Extreme problems")

survey_df_o <- survey_df_o[c(8, 3, 10, 5, 6, 1, 9, 4, 7, 2), ]
survey_df_y <- survey_df_y[c(8, 3, 10, 5, 6, 1, 9, 4, 7, 2), ]

survey_df_o <- tibble::rownames_to_column(survey_df_o, var="Domain")
survey_df_y <- tibble::rownames_to_column(survey_df_y, var="Domain")
survey_df_o <- survey_df_o[ , c(1,4,3,2)]
survey_df_y <- survey_df_y[ , c(1,4,3,2)]

origNames =  colnames(survey_df_o)

#resolve graphical superposition when one cell is 0 by adding a very low value to that cell

survey_df_o[7,4] <- 0.0001

likert(Domain ~., survey_df_o, as.percent=TRUE, 
       positive.order=TRUE, rightAxis=FALSE, 
       main = "EQ-5D domains - pre vs postoperative, \U2265 65 y",
       xlab="Percentage",
       panel=myPanelFunc)
```

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
origNames =  colnames(survey_df_y)

survey_df_y[2,2] <- 0.0001

likert(Domain ~., survey_df_y, as.percent=TRUE, 
       positive.order=TRUE, rightAxis=FALSE, 
       main = "EQ-5D domains - pre vs postoperative, <65 y",
       xlab="Percentage",
       panel=myPanelFunc)
```
  
---

**Model I - Pre-operative EQ5D predictors**
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}

library(sjPlot)
library(sjmisc)
library(sjlabelled)
library(magrittr)
library(knitr)

data_fact$Education3 <- relevel(data_fact$Education3, ref = 3)

mod_pre <-  lm(EQ5D_Pre~
                (Age + Symptoms_3m + Patologia2 + Gender + 
                   Education3 + Age * Education3),
               data=data_fact) 

#can not use tab_model and notebook, knit to html to see output
tab_model(mod_pre, show.se = TRUE, show.std = TRUE)
```
  
---

**Plotting Interaction Effects between Age and Education**  
**Preoperative QoL is related with education equal or less than 4 years**  

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

library(interactions)
library(jtools)

data_fact4 <- data_fact
data_fact4$Education3 <- 
  factor(data_fact4$Education3, 
         levels = c(1, 2,3), labels = c("educ. \U2264 4y","educ. 6-12y", ">12y"))
data_fact4$Education3 <- relevel(data_fact4$Education3, ref = 3)
mod_pre2 <-  lm(EQ5D_Pre~
                (Age + Symptoms_3m + Patologia2 + Gender + 
                   Education3 + Age * Education3),
               data=data_fact4) 

#the values for education used are <4y and 6-12y, we only have 2 patients with more than 12 y of educ. above 60 y old.



interact_plot(mod_pre2, pred = Age, modx = Education3, 
              modx.values = c("educ. \U2264 4y","educ. 6-12y"), 
              interval = TRUE, int.width = 0.8,
              main.title = "Interaction between age and education", 
              y.label = "Preoperative EQ5D", legend.main="Education level")  + theme_apa() + xlim(c(50,85))
```
  
---

**Model II - EQ5D improvement**

  
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
mod_pos2 <-  lm(Dif_EQ_pt~
                (Patologia2 + Psychiatric + Previous_Spine_Cx + Smoker + 
                   Symptoms_3m + Levels_Cx_number),
               data=data_fact) 
tab_model(mod_pos2, show.se = TRUE, show.std = TRUE)
```


